---
title: "Analysing differential hydrogen deuterium exchange mass spectrometry data"
author:
- name: Oliver M. Crook
package: hdxstats
output:
  BiocStyle::html_document:
    toc_float: yes
abstract: "This vignette describes how to analyse a mass-spectrometry based  hydrogen
  deuterium exchange experiment, in particular we focus on empirical Bayes functional
  models and visualisations. \n"
vignette: |
  %\VignetteIndexEntry{Analysing differential hydrogen deuterium exchange mass spectrometry data}
  %\VignetteEngine{knitr::rmarkdown}
  %%\VignetteKeywords{Mass Spectrometry, MS, MSMS, Proteomics, Metabolomics, Infrastructure, Quantitative} %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r env, message = FALSE, warning = FALSE, echo = FALSE}
library("hdxstats")
library("dplyr")
library("ggplot2")
library("RColorBrewer")
library("tidyr")
library("pheatmap")
library("scales")
library("viridis")
library("patchwork")
library("Biostrings")

```

# Analysis of a typical HDX-MS experiment

We have seen the basic aspects of our functional modelling approach. We now
wish to roll out our method across all peptides in the experiment. The
`fitUptakeKinetics` function allows us to apply our modelling approach across
all the peptide in the experiment. We need to provide a `QFeatures` object 
and the features for which we are fitting the model. The design will be extracted
from the column names or you can provide a design yourself. The parameter 
initilisation should also be provided. Sometimes the model can't be fit on the
kinetics. This is either because there is not enough data or through lack of 
convergence. An error will be reported in these cases but this should not 
perturb the user. You may wish to try a few starting values if there 
excessive models that fail fitting. 

```{r,}
res <- fitUptakeKinetics(object = MBPqDF[,c(1:24)],
                         feature = rownames(MBPqDF[,c(1:24)])[[1]],
                         start = list(a = NULL, b = 0.001,  d = NULL, p = 1))

```
The code chunk above returns a class `HdxStatModels` indicating that a number
of models for peptide have been fit. This is simply a holder for a list
of `HdxStatModel` instances.
```{r,}
res
```

We can easily examine indivual fits by going to the underyling `HdxStatModel`
class:
```{r,}
res@statmodels[[1]]@vis + scale_color_manual(values = brewer.pal(n = 2, name = "Set2"))
```
We now wish to apply statistical analysis to these fitted curves. Our approach
is an empirical Bayes testing procedure, which borrows information across peptides
to stablise variance estimates. Here, we need to provide the original data
that was analysed and the `HdxStatModels` class. The following code chunk
returns an object of class `HdxStatRes`. This object tell us that statistical
analysis was performed using our Functional model.

```{r,}
out <- processFunctional(object = MBPqDF[,1:24], params = res)
out
```

The main slot of interest is the `results` slot which returns quantities of 
interest such as `p-values` and `fdr` corrected p-values because of multiple testing.
The following is the `DataFrame` of interest.
```{r,}
out@results
```

We can now examine the peptides for which the false discovery rate is less
than 0.05

```{r,}
which(out@results$ebayes.fdr < 0.05)
```
Let us visualise some of these examples:
```{r,}
res@statmodels[[42]]@vis + res@statmodels[[45]]@vis

```
As we can see our model has picked up some subtle differences, we can further
visualise these using a forest plot. We can see the the functions are very similar
as the parameters are almost identical `(a,b,p,d)`. However, we can see that
the deuterium differences are lower in 10% structural variant condition.

```{r,}
fp <- forestPlot(params = res@statmodels[[42]])

```
We can produce a table to actual numbers. We see that at all 4 timepoints
the deuterium difference is negative, though the confidence intervals overlap
with 0. Our functional approach is picking up this small but reproducible difference.
```{r,}
knitr::kable(fp$data)
```

It is also possible to visualize, these plots on a different scale. Of course,
changing the natural scaling will emphasis different parts of the plot and
could distort interpretation. In particular, if a log transform is used then
care should be taken when interpreting values around 0. We suggest examining
the numerical values in a forest plot or table alongside any transformation of 
the variables. We suggest using the `pseudo log transform` as this allows
control the linearity of the plot, clearly demonstrating this a choice
of visualisation (and not of statistical modelling). The parameter `sigma`
below controls the scaling factor of the linear part of the transformation.

```{r,}
res@statmodels[[42]]@vis + scale_x_continuous(
    trans = pseudo_log_trans(base = 10, sigma = 0.01), breaks = c(0, 10^(1:7)))

res@statmodels[[42]]@vis + scale_x_continuous(
    trans = pseudo_log_trans(base = 10, sigma = 0.0001), breaks = c(0, 10^(1:7)))

res@statmodels[[42]]@vis + scale_x_continuous(
    trans = pseudo_log_trans(base = 10, sigma = 10), breaks = c(0, 10^(1:7)))

```



Let's us now have a look a situation where the changes are more dramatic.

```{r,}
res_wt <- fitUptakeKinetics(object = MBPqDF[, c(61:100)],
                            feature = rownames(MBPqDF[, c(61:100)])[[1]],
                            start = list(a = NULL, b = 0.001,  d = NULL, p = 1))

```
```{r,}
out_wt <- processFunctional(object = MBPqDF[, c(61:100)], params = res_wt)
```

We can visualise some of the result and generate plots.
```{r, fig.height = 16, fig.width = 20, fig.align = "center"}
res_wt@statmodels[[27]]@vis/res_wt@statmodels[[28]]@vis + plot_layout(guides = "collect")|(forestPlot(params = res_wt@statmodels[[27]], condition = c("WT", "W169G"))/forestPlot(params = res_wt@statmodels[[28]], condition = c("WT", "W169G")) + plot_layout(guides = "collect")) + 
    plot_annotation(tag_levels = 'a') +  plot_layout(widths = c(1, 1))
```
