---
title: "Analysing differential hydrogen deuterium exchange mass spectrometry data"
author:
- name: Oliver M. Crook
package: hdxstats
output:
  html_document:
    df_print: paged
  BiocStyle::html_document:
    toc_float: yes
abstract: "This vignette describes how to analyse a mass-spectrometry based  hydrogen
  deuterium exchange experiment, in particular we focus on empirical Bayes functional
  models and visualisations. \n"
vignette: |
  %\VignetteIndexEntry{Analysing differential hydrogen deuterium exchange mass spectrometry data} %\VignetteEngine{knitr::rmarkdown} %%\VignetteKeywords{Mass Spectrometry, MS, MSMS, Proteomics, Metabolomics, Infrastructure, Quantitative} %\VignetteEncoding{UTF-8}
---

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

```{r env, message = FALSE, warning = FALSE, echo = FALSE}
library("hdxstats")
```

# A well-defined HDX-MS experiment

This vignette describeds how to analyse time-resolved differential HDX-MS
experiments. The key elements are at least two conditions i.e. apo + antibody,
apo + small molecule or protein closed + protien open, etc. The experiment can
be replicated, though if there are sufficient time points analysed (>=3) then
occasionally signficant results can be obtained. The data provided should be
centroid-centric data. This package does not yet support analysis straight
from raw spectra. Typically this will be provided as a .csv from tools such as
dynamiX or HDExaminer.

# Main elements of the package

The package relies of Bioconductor infrastructure so that it integrates with
out datatypes and can benifit from advantages in other fields of mass-spectrometry.
There are package specific object, classes and methods but importantly there is
reuse of classes found in quantitative proteomics data, mainly the `QFeatures`
object which extends the `summarisedExperiment` class for mass spectrometry data.
The focus of this package is on testing and visualisation of the testing results.


